---
title: "Double-checking the MAST::zlm Result of the ASD Study with Permutations"
author: "Isabel Marleen Pötzsch, Simon Anders"
date: "2019-07-27"
output: html_document
---

#R setup

##Load Libraries and Data

Firstly, all relevant libraries, the raw data and the additional meta data are loaded. 
```{r}
knitr::opts_chunk$set(echo = TRUE)
#library( ggplot2 )
library( Matrix )
#library( locfit )
#library( purrr )
#library( furrr )
library( tidyverse )
library(MAST)
library(lme4)

#Load data
cellinfo <- read.delim( "~/tmp/ASD/meta.txt", stringsAsFactors=FALSE )
counts <- Seurat::Read10X("~/tmp/ASD/rawMatrix")
```


##Subset and Normalise Counts
Doublecheck whether the column names in the count matrix and the cell barcodes in the "meta.txt" file are the identical, then remove genes that do noty appear in any cell.
```{r}
stopifnot( identical( meta$cell, colnames(counts) ) )
counts <- counts[ rowSums(counts) > 0,  ]
```

We normalize by scaling to TPM, and taking log2 

```{r}
#Normalising counts
expr <- log1p(t( t(counts) / colSums(counts) )*1e6) / log(2)
```


##Scale Factors
To improve results using the zlm function, numeric factors are scaled and included in a data frame called `cellinfo_scaled`. This follows the code that Dmitry has sent us.
```{r}
#Scale factors for zlm
cellinfo_scaled <- data.frame(
  sample = cellinfo$sample,
  cluster = cellinfo$cluster,
  diagnosis = cellinfo$diagnosis,
  genes = scale(cellinfo$genes),
  age = scale(cellinfo$age),
  sex = cellinfo$sex,
  RIN = scale(cellinfo$RNA.Integrity.Number),
  PMI = scale(cellinfo$post.mortem.interval..hours.),
  region = cellinfo$region,
  Capbatch = cellinfo$Capbatch,
  Seqbatch = cellinfo$Seqbatch,
  ind = cellinfo$individual,
  ribo_perc = scale(cellinfo$RNA.ribosomal.percent))
```

#MAST analysis

##Create SCA object of cells in L2/3 cluster of one gene
For the MAST analysis, we create a SingleCellExperiment object first, including all cells in the L2/3 cluster, as given in cellinfo. This is converted into a SingleCellAssay object as required in the MAST package. This SingleCellAssay object is subset to only one gene to exemplify the work flow and cut down on computational demand.
```{r}
# SCE object
sce <- SingleCellExperiment( 
  assays = list( logTPM = expr ), 
  colData = cellinfo_scaled )

# Subset to only L2/3 cells
sce <- sce[ , sce$cluster == "L2/3" ]

# Convert to SCA object (required by MAST)
sca <- as( sce, 'SingleCellAssay' )

# Subsetting to a single gene only works if we convert the matrix to a dense one afterwards,
# so we write a function for that
subset_sca_to_gene <- function( sca, g ) {
  sca <- sca[ g, ]
  assay( sca ) <- as.matrix( assay( sca ) )
  return( sca )
}
```

##Fit ZLM

After creation of an appropriate SingleCellAssay object, a zero-inflated linear model is fitted via the zlm function from the MAST package and evaluated with a likelihood ratio test.

We test this for "TTF2", the most significant hit for L2/3:

```{r}
gene <- "TTF2"
```

```{r}
zlm_fit <- zlm( 
  ~ diagnosis + (1|ind) + genes + age + sex + RIN + PMI + region + Capbatch + Seqbatch + ribo_perc, 
  subset_sca_to_gene( sca, "TTF2" ), 
  method = "glmer", ebayes = F, silent=T, fitArgsD = list(nAGQ = 0))

#likelihood ratio test to obtain p-Values
lrTest( zlm_fit, Hypothesis( "diagnosisControl" ) )
```

We will use the value in the field labelled with "Pr(>Chisq)" and "hurdle" as the p value:

```{r}
lrTest( zlm_fit, Hypothesis( "diagnosisControl" ) )[ 1, "hurdle", "Pr(>Chisq)" ]
```

We hope that this is the same p value as used in the paper. (We cannot check as only the adjusted p values are given in the paper's supplement, and we cannot adjust without calculating for all genes.)

# Test validation by permutation

To validate the results seen in the zlm test, we use permutations: We randomly shuffle the assignments of diagnoses to the samples, then for each permutation, a zlm is fitted and a likelihood ratio test is performed. We run one hundred permutations and then check whether the p values from the permutations follow a uniform distribution, as they should nif the test were valid. 

## Do the permutation

First, get the assignment of samples to diagnoses

```{r}
cellinfo %>%
  select( sample, diagnosis ) %>%
  distinct ->
     diagn
head(diagn)
```

This can be easily shuffled:

```{r}
diagn %>% mutate( diagnosis_perm = sample(diagnosis) )
```
and this shuffled table can be easily added to the 'cellinfo' table as extra column

```{r}
sce_perm <- sce
colData(sce_perm)$diagnosis_perm <-
  left_join(         
     as.data.frame(colData(sca)),
     diagn %>% mutate( diagnosis_perm = sample(diagnosis) ) )$diagnosis_perm
```

With this, we can redo the zlm fit using the permuted column

```{r}
zlm_fit_perm <- MAST::zlm( 
  ~ diagnosis_perm + (1|ind) + genes + age + sex + RIN + PMI + region + Capbatch + Seqbatch + ribo_perc, 
  subset_sca_to_gene( as( sce_perm, "SingleCellAssay" ), "TTF2" ), 
  method = "glmer", ebayes = F, silent=T, fitArgsD = list(nAGQ = 0) )

lrTest( zlm_fit_perm, Hypothesis( "diagnosis_permControl" ) )[ 1, "hurdle", "Pr(>Chisq)" ]
```

# Do 100 permutations

We now do this 100 times
```{r}
n_perm <- 100   # number of permutations

pvalues_perm <- rep( NA, n_perm )
lambdas_perm <- rep( NA, n_perm )

for (i in 1:n_perm) {

  sce_perm <- sce
  colData(sce_perm)$diagnosis_perm <-
    left_join(         
      as.data.frame(colData(sca)),
      diagn %>% mutate( diagnosis_perm = sample(diagnosis) ) )$diagnosis_perm

  zlm_fit_perm <- MAST::zlm( 
    ~ diagnosis_perm + (1|ind) + genes + age + sex + RIN + PMI + region + Capbatch + Seqbatch + ribo_perc, 
    subset_sca_to_gene( as( sce_perm, "SingleCellAssay" ), "TTF2" ), 
    method = "glmer", ebayes = F, silent=T, fitArgsD = list(nAGQ = 0) )
  
  pvalues_perm[i] <-
    lrTest( zlm_fit_perm, Hypothesis( "diagnosis_permControl" ) )[ 1, "hurdle", "Pr(>Chisq)" ]
  lambdas_perm[i] <-
    lrTest( zlm_fit_perm, Hypothesis( "diagnosis_permControl" ) )[ 1, "hurdle", "lambda" ]

}
```
  
## Results

First, a histogram of the p values from the permutations:

```{r}
hist( pvalues_perm )
```
  
This looks still reasonable. However, the QQ plot does not:

The p-values generated for each of the permutations are sorted and the -log10 is plotted against the -log10 of a hundred evenly spaced points to generate a QQ plot.
```{r}
plot(
  -log10( ppoints( 100 ) ), 
  -log10( sort( pvalues_perm ) ), 
  main="QQ Plot", 
  xlab="-log10 of evenly-spaced points", 
  ylab="-log10 of p-values from MAST permutations" )
abline(0,1)
```

This QQ plot shows that the p-values generated by a random distribution of diagnoses using MAST are sytematically too small: there is a strong excess of very small p values, even though there should be no signal in the permutated data. This can also be seen by simply sorting them:

```{r}
sort( pvalues_perm )
```

Hence, it is possible that the significant values seen in the actual analysis are an artefact of the MAST package which does not seem to correctly control type-I error rate.

At least, none of the permuted test statistics (the "lambda" values) is larger than the actually observed statistic (31):

```{r}
sum( lambdas_perm > 31 )
```
Hence, the permutation p value is smaller than 1/100. We would need to do more permutations to get an actual value. It is unclear, however, that this value would survive adjustment for multiple testing.
